A fully labelled proof system for intuitionistic modal logics

### 2020-10-01 FullyLabelled/JLC/introduction.tex
up to section 6

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hypersequents~\cite{avron:elc96}, nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann:tableaux15},
labelled calculi have the advantage of being more uniform and being able to 
accommode a larger class of logics.

Standard labelled sequent calculi for logics with
standard Kripke semantics attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally use
\emph{relational atoms} of the form $\accs xy$ where 
the binary relation symbol $\rel$
%These calculi work best for logics with
%standard Kripke semantics, as in this case the relation $\rel$ 
is used to encode the accessibility relation in the Kripke models, such that
frame conditions corresponding to the desired logic can be directly
encoded as inference rules.
%\todo{Reviewer 1 complained about the ``best''}
Prominent examples are classical modal
logics and intuitionistic propositional logic, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$) can be translated into the
inference rule
\begin{equation}
\label{eq:Rtrans}
\vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas~\cite{negri:jpl2005}.

@@

	and considered as an inelegant solution~\cite{simpson:phd}. % (see also \cite{mar:str:tableaux17}).

@@

Besides that, there is another pleasant observation to make about our
system. It is naturally a multi-conclusion system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. 
%
The same is true for the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multi-conclusion systems in nested
sequents~\cite{str:2017maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:2014} for
intuitionistic logic.

A system similar to ours is presented
in~\cite{maffezioli:etal:synthese13}, for intuitionistic bimodal
epistemic logic. That system is also a multi-conclusion labelled system with
more than one relation. In fact, the logical rules are essentially the
same as ours. But we use a more general setting and provide deductive proof
systems for a larger class of intuitionistic modal logics, together
with an internal cut elimination procedure.

>>>>

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorists
to give sound, complete, and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hypersequents~\cite{avron:elc96}, nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 
2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann:tableaux15},
labelled calculi have the advantage of being more uniform and being able to 
accommode a larger class of logics.

Labelled sequent calculi for logics with
standard Kripke semantics attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally use
\emph{relational atoms} of the form $\accs xy$ where 
the binary relation symbol $\rel$
%These calculi work best for logics with
%standard Kripke semantics, as in this case the relation $\rel$ 
is used to encode the accessibility relation in the Kripke models, such that
frame conditions corresponding to the desired logic can be directly
encoded as inference rules.
%\todo{Reviewer 1 complained about the ``best''}
Prominent examples are classical modal
logics and intuitionistic propositional logic, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$) can be translated into the
inference rule
\begin{equation}
\label{eq:Rtrans}
\vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas~\cite{negri:jpl2005}.

@@

	and considered an inelegant solution~\cite{simpson:phd}. % (see also \cite{mar:str:tableaux17}).

@@

Besides that, there is another pleasant observation to make about our
system. It is naturally a multi-conclusion system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. 
%
The same is true for the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multi-conclusion systems in nested
sequents~\cite{str:2017maehara}, and therefore also in ordinary
labelled sequents. However, even in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:2014} for
intuitionistic logic.

A system similar to ours is presented
in~\cite{maffezioli:etal:synthese13}, for intuitionistic bimodal
epistemic logic. That system is also a multi-conclusion labelled system with
more than one relation. In fact, the logical rules are essentially the
same as ours, but we use a more general setting and provide deductive proof
systems for a larger class of intuitionistic modal logics, together
with an internal cut elimination procedure.

### 2020-10-01 FullyLabelled/JLC/preliminaries.tex
up to section 6

The idea is that intuitionistic propositional logic does not allow the principle of \emph{Excluded Middle}, so the modalities $\BOX$ and $\DIA$ are not de Morgan duals any more, but one can choose to design the axiomatisation in order to relate them in different ways. The most basic intuitionistic modal system one can think of would be to consider only the $\BOX$ modality as regulated by the $\kax$ axiom (or as called here $\kax[1]$), which gives the system \hbox{$\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$}. However this would give strictly no information on the behaviour of the $\DIA$ modality.
It seems that Fitch \cite{fitch:pm48} was the first one to propose a way to treat $\DIA$ in an intuitionistic system by considering the system  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$, which is now sometimes called $\rn{CK}$ for \emph{constructive modal logic}~\cite{bierman:depaiva:sl00,mendler:scheele:ic11}. Wijekesera \cite{wijesekera:apal90} also considered the axiom $\kax[5]$, which states that $\DIA$ distributes over 0-ary disjunctions, but did not assume that it would always distribute over binary disjunctions; the system he proposed was therefore  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$  $\rn{+}$ $\kax[5]$. In these systems, however, the addition of the \emph{Excluded Middle} principle to it does not yield classical modal logic $\K$, that is, it is not possible to retrieve the De Morgan duality of $\BOX$ and $\DIA$ in this case.

@@

The Kripke semantics for $\IK$ was first defined by Fischer-Servi~\cite{fischer-servi:84}.
%
It combines the Kripke semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.\footnote{This semantics and some variants have also been studied in~\cite{bovzic1984models}. For an excellent survey on possible variants see~\cite{simpson:phd}.}

\begin{definition}
	A \emph{bi-relational frame} $\F$ is a triple $\langle W, \rel, \le \rangle$ 
	%	of a non-empty set of worlds $W$ equipped with two binary relations $R$ and $\le$, where $R$ being the modal \emph{accessibility relation} and $\le$ a preorder (\emph{i.e.} a reflexive and transitive relation), satisfying the following conditions:
	of a set of worlds $W$ equipped with an {accessibility relation} $\rel$ and a preorder $\le$ (\emph{i.e.}\ a reflexive and transitive relation) satisfying:
	\begin{enumerate}
		\item[($\rn{F_1}$)] For all $\lb u, \lb v, \lb{v'} \in W$, if $\accs uv$ and $\futs v{v'}$, there exists $\lb{u'}$ s.t.~$\futs u{u'}$ and $\accs{u'}{v'}$.

		\begin{equation*}
		\xymatrix{
			\lb{u'} \ar@{.>}[r]^R  & \lb{v'} \\
			\lb u \ar@{.>}[u]^{\le} \ar@{->}[r]^R  & \lb v \ar@{->}[u]_{\le}
		}
		\end{equation*}

		\item[($\rn{F_2}$)] For all $\lb {u'}, \lb u, \lb v \in W$, if $\futs u{u'}$ and $\accs uv$, there exists $\lb {v'}$ s.t.~$\accs{u'}{v'}$ and $\futs v{v'}$.

		\begin{equation*}
		\xymatrix{
			\lb{u'} \ar@{.>}[r]^R & \lb{v'} \\
			\lb u \ar@{->}[u]^{\le} \ar@{->}[r]^R & \lb v \ar@{.>}[u]_{\le}
		}
		\end{equation*}
	\end{enumerate}
	%	
\end{definition}

>>>>

The idea is that intuitionistic propositional logic does not allow the principle of \emph{Excluded Middle}, so the modalities $\BOX$ and $\DIA$ are not de Morgan duals any more, but one can choose to design the axiomatisation in order to relate them in different ways. The most basic intuitionistic modal system one can think of would be to consider only the $\BOX$ modality as regulated by the $\kax$ axiom (or as called here $\kax[1]$), which gives the system \hbox{$\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$}. However this would give strictly no information on the behaviour of the $\DIA$ modality.
It seems that Fitch \cite{fitch:pm48} was the first one to propose a way to treat $\DIA$ in an intuitionistic system by considering the system  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$, which is now sometimes called $\rn{CK}$ for \emph{constructive modal logic}~\cite{bierman:depaiva:sl00,mendler:scheele:ic11}. Wijekesera \cite{wijesekera:apal90} also considered the axiom $\kax[5]$, which states that $\DIA$ distributes over 0-ary disjunctions, but did not assume that it would always distribute over binary disjunctions; the system he proposed was therefore  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$  $\rn{+}$ $\kax[5]$. In these systems, however, the addition of the \emph{Excluded Middle} does not yield classical modal logic $\K$, that is, it is not possible to retrieve the De Morgan duality of $\BOX$ and $\DIA$ in this case.

@@

The Kripke semantics for $\IK$ was first defined by Fischer-Servi~\cite{fischer-servi:84}.
%
It combines the semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.\footnote{This semantics and some variants have also been studied in~\cite{bovzic1984models}. For an excellent survey on possible variants see~\cite{simpson:phd}.}

\begin{definition}
	A \emph{bi-relational frame} $\F$ is a triple $\langle W, \rel, \le \rangle$ 
	%	of a non-empty set of worlds $W$ equipped with two binary relations $R$ and $\le$, where $R$ being the modal \emph{accessibility relation} and $\le$ a preorder (\emph{i.e.} a reflexive and transitive relation), satisfying the following conditions:
	of a set of worlds $W$ equipped with an {accessibility relation} $\rel$ and a preorder $\le$ (\emph{i.e.}\ a reflexive and transitive relation) satisfying:
	\begin{enumerate}
		\item[($\rn{F_1}$)] For all $\lb x, \lb y, \lb z \in W$, if $\accs xy$ and $\futs yz$, there exists $\lb u$ s.t.~$\futs xu$ and $\accs uz$.

		\begin{equation*}
		\xymatrix{
			\lb u \ar@{.>}[r]^R  & \lb z \\
			\lb x \ar@{.>}[u]^{\le} \ar@{->}[r]^R  & \lb y \ar@{->}[u]_{\le}
		}
		\end{equation*}

		\item[($\rn{F_2}$)] For all $\lb x, \lb y, \lb z \in W$, if $\futs xz$ and $\accs xy$, there exists $\lb u$ s.t.~$\accs zu$ and $\futs yu$.

		\begin{equation*}
		\xymatrix{
			\lb z\ar@{.>}[r]^R & \lb u \\
			\lb x \ar@{->}[u]^{\le} \ar@{->}[r]^R & \lb y \ar@{.>}[u]_{\le}
		}
		\end{equation*}
	\end{enumerate}
	%	
\end{definition}

### 2020-10-01 FullyLabelled/JLC/soundness.tex
up to section 6

\begin{proof}
	We proceed by induction on the height of the derivation of $\SG$, and we show for all rules in $\labIKp$
	$$
	\vliiinf{\labrn{r}}{}{\SG}{\SGi1}{\cdots}{\SGi n}
	$$ for $n\in\{0,1,2\}$, that whenever $\SGi1,\ldots,\SGi n$ are
	valid in all birelational frame, then so is $\SG$. It follows a case analysis on $\rn r$:
	\begin{itemize}
		\item $\rlabrn\bot$: This is trivial because $\bot$ is never forced.
		\item $\rn{id}$: This follows immediately from Proposition~\ref{prop:mon-adm}.
		\item $\llabrn\BOX$: By way of contradiction, assume that $\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right$
		is valid in all birelational frames, but $\B, \futs xy,	\accs yz,\Left, \labels{x}{\BOX A} \SEQ \Right$ is not. This means that
		we have a model $\M$ and an interpretation $\inter\cdot$, such
		that $\M,\inter\cdot\not\Vdash\B, \futs xy, \accs yz, \Left,	\labels{x}{\BOX A} \SEQ \Right$, i.e., $\inter x\le_\M\inter y$
		and $\inter y\rel_\M\inter z$ and $\cforce \M x{\BOX A}$ and
		$\cnforce \M wB$ for all $\labels wB\in\Right$. However, by the
		definition of forcing in~\eqref{eq:kripke} we also have $\cforce
		\M zA$, and consequently $\nforce \M{\inter\cdot}{\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ
			\Right}$. Contradiction.
%		\todo{The reasoning by contradiction is not necessary. Could be rewritten more lightly as follows:}
%			%
%			Assume that $\B, \Left, \futs xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right$
%			is valid in all birelational frames. 
%			Consider a model $\M$ and an interpretation $\inter\cdot$ of the conclusion sequent
%			$\B,\Left, \futs xy, \accs yz, \labels{x}{\BOX A} \SEQ \Right$ into $\M$, such
%			that for all $\labels{z}{C} \in \Gamma$ we have $\M, \inter z \Vdash \fm C$
%			and $\cforce \M x{\BOX A}$. %and $\cnforce \M wB$ for all $\labels wB\in\Right$. However, 
%			Hence, as $\inter x\le_\M\inter y$ and $\inter y\rel_\M\inter z$, 
%			by the definition of forcing in~\eqref{eq:kripke} we also have $\cforce \M zA$, and 
%			consequently $\cforce \M wB$ for all $\labels wB\in\Right$. 
%			Thus, $\cforce \M{\inter\cdot}{\B, \Left, \futs
%				xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \SEQ
%				\Right}$. \todo{should we rewrite the whole proof?}\lutz{no. we do by contradiction. I'll explainon skype.}

>>>>

\begin{proof}
	We proceed by induction on the height of the derivation of $\SG$, and we show for all rules in $\labIKp$
	$$
	\vliiinf{\labrn{r}}{}{\SG}{\SGi1}{\cdots}{\SGi n}
	$$ for $n\in\{0,1,2\}$, that whenever $\SGi1,\ldots,\SGi n$ are
	valid in all birelational frame, then so is $\SG$. It follows from a case analysis on $\rn r$:
	\begin{itemize}
		\item $\rlabrn\bot$: This is trivial because $\bot$ is never forced.
		\item $\rn{id}$: This follows immediately from Proposition~\ref{prop:mon-adm}.
		\item $\llabrn\BOX$: By way of contradiction, assume that $\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right$
		is valid in all birelational frames, but $\B, \futs xy,	\accs yz,\Left, \labels{x}{\BOX A} \SEQ \Right$ is not. This means that
		we have a model $\M$ and an interpretation $\inter\cdot$, such
		that $\M,\inter\cdot\not\Vdash\B, \futs xy, \accs yz, \Left,	\labels{x}{\BOX A} \SEQ \Right$, i.e., $\inter x\le_\M\inter y$
		and $\inter y\rel_\M\inter z$ and $\cforce \M x{\BOX A}$ but
		$\cnforce \M wB$ for all $\labels wB\in\Right$. However, by the
		definition of forcing in~\eqref{eq:kripke} we also have $\cforce
		\M zA$, and consequently $\nforce \M{\inter\cdot}{\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ
			\Right}$. Contradiction.
%		\todo{The reasoning by contradiction is not necessary. Could be rewritten more lightly as follows:}
%			%
%			Assume that $\B, \Left, \futs xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right$
%			is valid in all birelational frames. 
%			Consider a model $\M$ and an interpretation $\inter\cdot$ of the conclusion sequent
%			$\B,\Left, \futs xy, \accs yz, \labels{x}{\BOX A} \SEQ \Right$ into $\M$, such
%			that for all $\labels{z}{C} \in \Gamma$ we have $\M, \inter z \Vdash \fm C$
%			and $\cforce \M x{\BOX A}$. %and $\cnforce \M wB$ for all $\labels wB\in\Right$. However, 
%			Hence, as $\inter x\le_\M\inter y$ and $\inter y\rel_\M\inter z$, 
%			by the definition of forcing in~\eqref{eq:kripke} we also have $\cforce \M zA$, and 
%			consequently $\cforce \M wB$ for all $\labels wB\in\Right$. 
%			Thus, $\cforce \M{\inter\cdot}{\B, \Left, \futs
%				xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \SEQ
%				\Right}$. \todo{should we rewrite the whole proof?}\lutz{no. we do by contradiction. I'll explainon skype.}
